# Multivariate Analysis {#sec-multivariate}

```{r}
#| label: setup-ch13
#| include: false
library(tidyverse)
library(factoextra)
library(cluster)
set.seed(42)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Conduct and interpret Principal Component Analysis (PCA)
- Choose the number of components using a scree plot
- Perform K-means and hierarchical clustering
- Evaluate cluster quality and choose the optimal number of clusters
- Visualise multivariate results with `ggplot2` and `factoextra`
:::

## Principal Component Analysis {#sec-pca}

**PCA** reduces a high-dimensional dataset to a smaller set of uncorrelated **principal components** that capture the maximum variance in the data.

```{r}
#| label: pca-data

# Use USArrests dataset — violent crime rates in US states
data(USArrests)
head(USArrests)
```

```{r}
#| label: pca-fit

# Scale variables before PCA (important when variables have different units)
pca_result <- prcomp(USArrests, scale. = TRUE, center = TRUE)

# Variance explained
summary(pca_result)
```

```{r}
#| label: fig-scree
#| fig-cap: "Scree plot: the first two components explain ~62% of total variance."
#| fig-width: 7
#| fig-height: 4.5

fviz_eig(pca_result,
         addlabels = TRUE,
         barfill    = "#3498db",
         barcolor   = "white",
         linecolor  = "#e74c3c",
         ggtheme    = theme_minimal(base_size = 12)) +
  labs(title = "Scree Plot: Variance Explained by Each PC")
```

```{r}
#| label: fig-pca-biplot
#| fig-cap: "PCA biplot: states projected onto PC1 and PC2, with variable loadings as arrows."
#| fig-width: 8
#| fig-height: 6

fviz_pca_biplot(
  pca_result,
  repel      = TRUE,
  col.var    = "#e74c3c",
  col.ind    = "#2c3e50",
  label      = "all",
  ggtheme    = theme_minimal(base_size = 11)
) +
  labs(title = "PCA Biplot: US States by Crime Profile")
```

### Interpreting PCA {#sec-pca-interpret}

```{r}
#| label: pca-loadings

# Variable loadings (how much each variable contributes to each PC)
pca_result$rotation

# State scores on each component
head(pca_result$x)

# Proportion of variance
prop_var <- (pca_result$sdev^2) / sum(pca_result$sdev^2)
cumsum(prop_var)
```

## K-Means Clustering {#sec-kmeans}

K-means partitions observations into K groups by minimising within-cluster sum of squares.

```{r}
#| label: kmeans-data

# Scale the data first
arrests_scaled <- scale(USArrests)
```

### Choosing K {#sec-choosing-k}

```{r}
#| label: fig-elbow
#| fig-cap: "Elbow plot to choose the optimal number of clusters."
#| fig-width: 7
#| fig-height: 4

fviz_nbclust(arrests_scaled, kmeans, method = "wss") +
  geom_vline(xintercept = 4, linetype = "dashed", colour = "#e74c3c") +
  labs(title = "Elbow Method for Optimal K",
       x = "Number of Clusters", y = "Total Within SS") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: fig-silhouette
#| fig-cap: "Average silhouette width — K=4 appears optimal."
#| fig-width: 7
#| fig-height: 4

fviz_nbclust(arrests_scaled, kmeans, method = "silhouette") +
  labs(title = "Silhouette Method") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: kmeans-fit

set.seed(42)
km_result <- kmeans(arrests_scaled, centers = 4, nstart = 25)

# Cluster sizes
table(km_result$cluster)

# Cluster centres (in original scale)
km_result$centers
```

```{r}
#| label: fig-kmeans-plot
#| fig-cap: "K-means clustering of US states on PCA components."
#| fig-width: 8
#| fig-height: 6

fviz_cluster(
  km_result,
  data        = arrests_scaled,
  palette     = "Set2",
  ellipse.type = "convex",
  repel       = TRUE,
  ggtheme     = theme_minimal(base_size = 11)
) +
  labs(title = "K-Means Clusters (K=4): US States by Crime Profile")
```

## Hierarchical Clustering {#sec-hierarchical}

Hierarchical clustering builds a **dendrogram** — a tree that shows how observations merge into clusters at different distances.

```{r}
#| label: hclust-fit

# Compute distance matrix
dist_matrix <- dist(arrests_scaled, method = "euclidean")

# Hierarchical clustering with complete linkage
hclust_result <- hclust(dist_matrix, method = "complete")
```

```{r}
#| label: fig-dendrogram
#| fig-cap: "Dendrogram of US states. Cut at K=4 to match K-means."
#| fig-width: 10
#| fig-height: 6

fviz_dend(
  hclust_result,
  k           = 4,
  k_colors    = RColorBrewer::brewer.pal(4, "Set2"),
  rect        = TRUE,
  label_cols  = "black",
  cex         = 0.7,
  ggtheme     = theme_minimal(base_size = 10)
) +
  labs(title = "Hierarchical Clustering Dendrogram (Complete Linkage)")
```

```{r}
#| label: hclust-groups

# Cut the tree to get 4 clusters
hclust_groups <- cutree(hclust_result, k = 4)
table(hclust_groups)

# Compare K-means vs. hierarchical
table("K-means" = km_result$cluster, "H-clust" = hclust_groups)
```

## Exercises {#sec-ch13-exercises}

1. Apply PCA to the `mtcars` dataset. How many components are needed to explain 80% of variance? Create a biplot and describe which cars cluster together.

2. Use K-means clustering (choose K using the elbow method) to group countries in the `gapminder` 2007 dataset using `gdpPercap` and `lifeExp`. Map the clusters using a scatter plot coloured by cluster.

3. Compare K-means and hierarchical clustering on the `iris` dataset. How well do the clusters recover the true species labels? Compute accuracy.

4. **Challenge:** Apply PCA to India's district-level agricultural data (multiple crop yields as variables). Create a map (using `sf` package) coloured by cluster membership to show agricultural zones.
