# Time Series Analysis {#sec-timeseries}

```{r}
#| label: setup-ch14
#| include: false
library(tidyverse)
library(forecast)
library(tsibble)
library(feasts)
set.seed(42)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Create and manipulate time series objects in R
- Decompose a time series into trend, seasonal, and remainder components
- Test for stationarity using the ADF test
- Fit ARIMA models using `auto.arima()`
- Produce and visualise forecasts with confidence intervals
:::

## Time Series Objects {#sec-ts-objects}

```{r}
#| label: ts-create

# Base R ts object: monthly retail sales (simulated)
set.seed(99)
monthly_sales <- ts(
  data      = round(1000 + seq(0, 200, length.out = 60) +    # Trend
                    100 * sin(2 * pi * (1:60) / 12) +         # Seasonality
                    rnorm(60, 0, 40)),                         # Noise
  start     = c(2019, 1),    # January 2019
  frequency = 12             # Monthly
)

print(monthly_sales)
```

```{r}
#| label: fig-ts-plot
#| fig-cap: "Monthly retail sales time series with trend and seasonal patterns."
#| fig-width: 9
#| fig-height: 4

autoplot(monthly_sales) +
  labs(title = "Monthly Retail Sales (Simulated)",
       x = "Year", y = "Sales (₹ thousands)") +
  theme_minimal(base_size = 13)
```

## Time Series Decomposition {#sec-decomposition}

### Classical Decomposition {#sec-classical-decomp}

```{r}
#| label: fig-decomp
#| fig-cap: "STL decomposition separating trend, seasonality, and remainder."
#| fig-width: 9
#| fig-height: 7

decomp <- stl(monthly_sales, s.window = "periodic")

autoplot(decomp) +
  labs(title = "STL Decomposition of Monthly Sales") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: decomp-extract

# Extract components
trend     <- decomp$time.series[, "trend"]
seasonal  <- decomp$time.series[, "seasonal"]
remainder <- decomp$time.series[, "remainder"]

# Seasonal strength
var(seasonal) / (var(seasonal) + var(remainder))
```

## Stationarity {#sec-stationarity}

Most time series models require **stationarity** — constant mean, variance, and autocovariance over time.

```{r}
#| label: fig-acf-pacf
#| fig-cap: "ACF and PACF plots help identify appropriate ARIMA orders."
#| fig-width: 9
#| fig-height: 5

par(mfrow = c(1, 2))
acf(monthly_sales, main = "ACF: Monthly Sales")
pacf(monthly_sales, main = "PACF: Monthly Sales")
par(mfrow = c(1, 1))
```

```{r}
#| label: stationarity-test

library(tseries)

# Augmented Dickey-Fuller test
# H₀: series has a unit root (non-stationary)
# H₁: series is stationary
adf.test(monthly_sales)

# If non-stationary, differencing can help
adf.test(diff(monthly_sales))
```

## ARIMA Models {#sec-arima}

**ARIMA(p, d, q)** models combine:
- **AR(p):** Autoregressive — depends on p past values
- **I(d):** Integrated — d-th order differencing for stationarity
- **MA(q):** Moving average — depends on q past errors

For seasonal data, use **SARIMA(p, d, q)(P, D, Q)[m]**.

```{r}
#| label: arima-fit

# auto.arima selects the best model automatically
m_arima <- auto.arima(
  monthly_sales,
  seasonal     = TRUE,
  stepwise     = FALSE,
  approximation = FALSE
)

summary(m_arima)
```

```{r}
#| label: fig-arima-diag
#| fig-cap: "ARIMA model residual diagnostics."
#| fig-width: 9
#| fig-height: 6

checkresiduals(m_arima)
```

## Forecasting {#sec-forecasting}

```{r}
#| label: fig-forecast
#| fig-cap: "12-month ahead forecast with 80% and 95% prediction intervals."
#| fig-width: 9
#| fig-height: 5

forecast_result <- forecast(m_arima, h = 12)

autoplot(forecast_result) +
  labs(title = "Retail Sales: 12-Month Forecast",
       x = "Year", y = "Sales (₹ thousands)",
       caption = "Shaded regions: 80% and 95% prediction intervals") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: forecast-accuracy

# Hold-out validation: train on first 48 months, test on last 12
train  <- head(monthly_sales, 48)
test   <- tail(monthly_sales, 12)

m_train <- auto.arima(train, seasonal = TRUE)
fc      <- forecast(m_train, h = 12)

# Accuracy metrics
accuracy(fc, test)
```

## The tsibble Framework {#sec-tsibble}

For modern, tidy time series workflows:

```{r}
#| label: tsibble-demo
library(tsibble)
library(feasts)

# Convert to tsibble
sales_tsibble <- as_tsibble(monthly_sales)

# Decompose
sales_tsibble |>
  model(STL(value ~ trend(window = 13) + season(window = "periodic"))) |>
  components() |>
  autoplot() +
  theme_minimal(base_size = 11) +
  labs(title = "STL Decomposition (tsibble/feasts)")
```

## Exercises {#sec-ch14-exercises}

1. Download monthly diesel or petrol price data for India from the PPAC website. Import and convert to a `ts` object. Plot the series and describe the trend and any seasonality.

2. Decompose the price series using STL. Plot the components and estimate the seasonal strength.

3. Fit an ARIMA model using `auto.arima()`. Print the selected order and the model summary.

4. Produce a 24-month forecast. Visualise it and write a brief interpretation of the projection.

5. **Challenge:** Build a comparison of three forecasting methods — Exponential Smoothing, ARIMA, and a linear trend model — evaluated by out-of-sample RMSE on a holdout set of 12 months.
