# Your First Reproducible Document with Quarto {#sec-quarto}

```{r}
#| label: setup-ch2
#| include: false
library(tidyverse)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Explain what Quarto is and how it differs from R Markdown
- Describe the anatomy of a `.qmd` file
- Use YAML headers to control document metadata and output options
- Control code chunk behaviour using chunk options
- Render a document to HTML, PDF, and Word
- Create a simple reproducible report from a real dataset
:::

## Why Quarto? {#sec-why-quarto}

For many years, R users relied on **R Markdown** (`.Rmd`) to write reproducible documents. Quarto is its successor — created by Posit (formerly RStudio) to be language-agnostic (it works with R, Python, Julia, and Observable JS) and to consolidate years of lessons learned from R Markdown.

The key conceptual shift is this: **your document and your analysis are the same file**. You do not write your analysis in R, copy the output, and paste it into a Word document. You write prose and code together in one `.qmd` file, and Quarto renders the whole thing into a polished output.

::: {.callout-tip}
## Quarto vs. R Markdown

If you are coming from R Markdown, here are the key differences:

| Feature | R Markdown | Quarto |
|---------|-----------|--------|
| File extension | `.Rmd` | `.qmd` |
| Chunk options | In `{}` header | YAML-style `#\|` comments |
| Language support | Primarily R | R, Python, Julia, Observable |
| Publishing | Various packages | Built-in (`quarto publish`) |
| Books/websites | `bookdown`, `distill` | Built-in project types |

Quarto can even render old `.Rmd` files without changes.
:::

## Anatomy of a `.qmd` File {#sec-anatomy}

Every Quarto document has three components: a **YAML header**, **Markdown prose**, and **code chunks**.

### The YAML Header {#sec-yaml}

The document begins with a **YAML header** delimited by triple dashes (`---`). It controls the document's metadata and output format.

```yaml
---
title: "My First Report"
author: "Pawan Kumar"
date: last-modified
format:
  html:
    toc: true
    code-fold: true
  pdf:
    toc: true
---
```

YAML is whitespace-sensitive. Indentation matters. Common YAML options include:

```{r}
#| label: tbl-yaml-options
#| echo: false
#| tbl-cap: "Common Quarto YAML options"

yaml_opts <- data.frame(
  Option = c("title", "author", "date", "format", "execute", "bibliography", "toc", "number-sections"),
  Purpose = c(
    "Document title",
    "Author name(s)",
    "Publication date (use `last-modified` to auto-update)",
    "Output format (html, pdf, docx)",
    "Default code execution options",
    "Path to .bib file for citations",
    "Include table of contents",
    "Number section headings"
  )
)
knitr::kable(yaml_opts)
```

### Markdown Prose {#sec-markdown}

Everything outside of code chunks is **Markdown** — a lightweight markup language that uses plain-text symbols to indicate formatting.

```markdown
# Heading 1
## Heading 2
### Heading 3

**Bold text**
*Italic text*
`inline code`

- Bullet point 1
- Bullet point 2
  - Nested bullet

1. Numbered item
2. Numbered item

[Link text](https://example.com)

> This is a blockquote.

| Col 1 | Col 2 |
|-------|-------|
| A     | 1     |
| B     | 2     |
```

### Code Chunks {#sec-chunks}

Code chunks are enclosed in triple backticks with `{r}` to specify the language:

````markdown
```{{r}}
#| label: my-first-chunk
#| echo: true
#| eval: true

mean(c(10, 20, 30, 40))
```
````

The `#|` prefix marks **chunk options** — YAML-style directives that control how the chunk behaves.

## Code Chunk Options {#sec-chunk-options}

Chunk options are the most powerful way to control your document's output. They are written as YAML comments at the top of each chunk:

```{r}
#| label: tbl-chunk-options
#| echo: false
#| tbl-cap: "Most important Quarto chunk options"

chunk_opts <- data.frame(
  Option = c("label", "echo", "eval", "include", "warning", "message",
             "output", "fig-cap", "fig-width", "fig-height", "cache"),
  Type = c("string", "logical", "logical", "logical", "logical", "logical",
           "logical", "string", "number", "number", "logical"),
  Default = c("none", "true", "true", "true", "true", "true",
              "true", "none", "7", "5", "false"),
  Effect = c(
    "Unique chunk identifier — required for cross-references",
    "Show the code in the output?",
    "Run the code?",
    "Include chunk (code + output) in document?",
    "Show R warnings?",
    "Show R messages?",
    "Show printed output (text)?",
    "Figure caption",
    "Figure width in inches",
    "Figure height in inches",
    "Cache chunk results to disk"
  )
)
knitr::kable(chunk_opts, col.names = c("Option", "Type", "Default", "Effect"))
```

### Demonstration: Controlling Output

Here is the same code with different chunk options to show the effect:

**Show code and output (`echo: true`, `output: true`):**

```{r}
#| label: demo-echo-true
#| echo: true
summary(mtcars$mpg)
```

**Show output only (`echo: false`):**

```{r}
#| label: demo-echo-false
#| echo: false
cat("This code was hidden, but it ran. Mean MPG:", round(mean(mtcars$mpg), 2))
```

**Show code only, do not run (`eval: false`):**

```{r}
#| label: demo-eval-false
#| eval: false
# This code block will not be executed
very_slow_function_that_takes_hours()
```

**Neither shown nor run (`include: false`):**

```{r}
#| label: demo-include-false
#| include: false
# This chunk is completely invisible in the output
x <- 42
```

The variable `x` was set to `r x` in the hidden chunk above — we can use it inline!

### Global Options with `execute`

Rather than setting options in every chunk, set defaults in the YAML header:

```yaml
execute:
  echo: true
  warning: false
  message: false
  cache: false
```

Individual chunks can then override these defaults as needed.

## Inline Code {#sec-inline}

One of the most powerful features of Quarto is **inline code** — R expressions embedded directly in your prose using backticks:

```markdown
The dataset contains `` `r nrow(mtcars)` `` rows and 
`` `r ncol(mtcars)` `` columns. The mean fuel efficiency 
is `` `r round(mean(mtcars$mpg), 1)` `` miles per gallon.
```

This renders as:

The dataset contains `r nrow(mtcars)` rows and `r ncol(mtcars)` columns. The mean fuel efficiency is `r round(mean(mtcars$mpg), 1)` miles per gallon.

::: {.callout-important}
## Never Hard-Code Numbers in Your Reports

If your dataset changes, inline code updates automatically. Hard-coded numbers go stale and introduce errors. Always compute values from code.
:::

## Rendering Documents {#sec-rendering}

### From RStudio

Click the **Render** button (or press `Ctrl+Shift+K`). RStudio will render to the format specified in the YAML header.

### From the Terminal

```bash
# Render to the default format
quarto render document.qmd

# Render to a specific format
quarto render document.qmd --to html
quarto render document.qmd --to pdf
quarto render document.qmd --to docx

# Render all files in a project
quarto render
```

### Multiple Output Formats

To render to multiple formats simultaneously, specify them all in the YAML:

```yaml
format:
  html:
    toc: true
  pdf:
    toc: true
  docx:
    reference-doc: template.docx
```

## Cross-References {#sec-crossref}

Quarto has a built-in cross-reference system. Label a figure or table, then reference it from anywhere in the document.

**Figures:**

````markdown
```{{r}}
#| label: fig-mpg-hist
#| fig-cap: "Distribution of fuel efficiency in the mtcars dataset"

hist(mtcars$mpg, main = "", xlab = "Miles per Gallon")
```
````

Then reference it as: `As shown in @fig-mpg-hist...`

```{r}
#| label: fig-mpg-hist
#| fig-cap: "Distribution of fuel efficiency in the mtcars dataset."
#| fig-alt: "Histogram showing distribution of MPG values"

hist(mtcars$mpg,
     main = "",
     xlab = "Miles per Gallon",
     col  = "#3498db",
     border = "white")
```

As shown in @fig-mpg-hist, fuel efficiency follows an approximately normal distribution with a slight right skew.

**Tables:**

````markdown
```{{r}}
#| label: tbl-mtcars-summary
#| tbl-cap: "Summary statistics for mtcars"

knitr::kable(summary(mtcars[, 1:4]))
```
````

Reference as: `@tbl-mtcars-summary shows...`

## Callout Blocks {#sec-callouts}

Callout blocks draw attention to important information:

```markdown
::: {.callout-note}
## Note Title
This is a note.
:::

::: {.callout-tip}
## Tip Title
This is a tip.
:::

::: {.callout-warning}
## Warning Title
This is a warning.
:::

::: {.callout-important}
## Important Title
This is important.
:::

::: {.callout-caution}
## Caution Title
Proceed with caution.
:::
```

## Hands-On Exercise: Your First Report {#sec-first-report}

Let us create a complete mini-report using the `mtcars` dataset. This exercise brings together everything in this chapter.

Create a new file `my-first-report.qmd` and add the following:

```yaml
---
title: "Motor Trend Cars: A Brief Analysis"
author: "Your Name"
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---
```

Then add these sections with code:

```{r}
#| label: first-report-setup

# Load the data
data(mtcars)

# Add row names as a column
mtcars_df <- mtcars |>
  tibble::rownames_to_column("car") |>
  tibble::as_tibble()

# How large is the dataset?
n_cars <- nrow(mtcars_df)
n_vars <- ncol(mtcars_df)
```

The `mtcars` dataset contains data on `r n_cars` cars and `r n_vars` variables, extracted from the 1974 *Motor Trend* magazine.

```{r}
#| label: tbl-mtcars-head
#| tbl-cap: "First six rows of the mtcars dataset"

knitr::kable(head(mtcars_df[, 1:7]))
```

```{r}
#| label: fig-mpg-cyl
#| fig-cap: "Fuel efficiency varies substantially by number of cylinders."

library(ggplot2)

ggplot(mtcars_df, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x     = "Number of Cylinders",
    y     = "Miles per Gallon",
    fill  = "Cylinders",
    title = "Fuel Efficiency by Number of Cylinders"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

@fig-mpg-cyl shows that cars with 4 cylinders achieve substantially better fuel efficiency than 6- or 8-cylinder cars.

::: {.callout-note}
## Try It Yourself

Extend this report by adding a scatter plot of `mpg` vs `wt` (weight). Add a linear regression line using `geom_smooth(method = "lm")`. What does the relationship tell you?
:::

## Publishing Your Document {#sec-publishing}

Once you have a document you are proud of, Quarto makes publishing straightforward:

```bash
# Publish to Quarto Pub (free hosting)
quarto publish quarto-pub document.qmd

# Publish to GitHub Pages
quarto publish gh-pages

# Publish to Netlify
quarto publish netlify
```

For RStudio users: *File → Publish Document* provides a GUI for publishing to RPubs or Connect.

## Exercises {#sec-ch2-exercises}

1. Create a new `.qmd` file. Write a YAML header that renders to both HTML and PDF. Set `warning: false` and `message: false` globally.

2. Load the `airquality` dataset. Create a report with three sections: (a) an overview of the data using `str()` and inline code reporting the number of rows; (b) a histogram of ozone levels; (c) a cross-reference to that histogram in your prose.

3. Demonstrate the difference between `echo: false` and `include: false` in the same document. Explain what you see in the output.

4. Use a callout block to highlight one important finding from your `airquality` analysis.

5. **Challenge:** Parameterise your report so that the variable being analysed (ozone, wind, or temperature) can be changed by modifying a single value at the top of the YAML. (Hint: look up Quarto parameters with `params:`.)
