# Writing Efficient R Code {#sec-efficient-r}

```{r}
#| label: setup-ch15
#| include: false
library(tidyverse)
library(purrr)
library(microbenchmark)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Use control structures: if-else, for loops, while loops
- Apply the `apply` family of functions for vectorised operations
- Use `purrr::map()` for functional programming
- Profile code performance and identify bottlenecks
- Write faster R code using vectorisation
:::

## Control Structures {#sec-control-structures}

### If-Else Statements {#sec-ifelse}

```{r}
#| label: ifelse-demo

# Basic if-else
classify_bmi <- function(bmi) {
  if (bmi < 18.5) {
    "Underweight"
  } else if (bmi < 25.0) {
    "Normal"
  } else if (bmi < 30.0) {
    "Overweight"
  } else {
    "Obese"
  }
}

classify_bmi(22.3)
classify_bmi(31.5)

# Vectorised version using dplyr::case_when
bmi_values <- c(17.2, 22.3, 27.8, 33.1, 19.5)

dplyr::case_when(
  bmi_values < 18.5 ~ "Underweight",
  bmi_values < 25.0 ~ "Normal",
  bmi_values < 30.0 ~ "Overweight",
  TRUE              ~ "Obese"
)
```

::: {.callout-tip}
## Prefer Vectorised Operations

In R, applying a function to each element of a vector using a loop is almost always slower than using vectorised operations. Use `ifelse()` or `case_when()` instead of `if` inside a loop.
:::

### For Loops {#sec-for-loops}

```{r}
#| label: for-loops

# A for loop to read and process multiple files
# (demonstrated conceptually)

results <- list()
for (i in 1:5) {
  results[[i]] <- tibble(
    simulation = i,
    mean_val   = mean(rnorm(100, mean = i * 10)),
    sd_val     = sd(rnorm(100))
  )
}

bind_rows(results)
```

### While Loops {#sec-while-loops}

```{r}
#| label: while-loops

# Newton-Raphson: find root of f(x) = x^2 - 2 (i.e., sqrt(2))
x    <- 1.0
tol  <- 1e-10
iter <- 0

while (abs(x^2 - 2) > tol) {
  x    <- x - (x^2 - 2) / (2 * x)
  iter <- iter + 1
}

cat(sprintf("sqrt(2) ≈ %.10f (converged in %d iterations)\n", x, iter))
```

## The Apply Family {#sec-apply-family}

The `apply` family avoids explicit loops and is often more readable (though not always faster):

```{r}
#| label: apply-demo

# apply: applies a function over rows (MARGIN=1) or columns (MARGIN=2)
mat <- matrix(rnorm(20), nrow = 4)
apply(mat, MARGIN = 1, mean)   # Row means
apply(mat, MARGIN = 2, sd)     # Column SDs

# lapply: applies to each element of a list, returns a list
datasets <- list(mtcars = mtcars, iris = iris, airquality = airquality)
lapply(datasets, nrow)

# sapply: like lapply but simplifies the result
sapply(datasets, nrow)          # Returns named vector
sapply(datasets, function(d) c(rows = nrow(d), cols = ncol(d)))

# vapply: like sapply but type-safe (recommended for production code)
vapply(datasets, nrow, FUN.VALUE = integer(1))

# tapply: apply by group
tapply(iris$Sepal.Length, iris$Species, mean)
```

## Functional Programming with `purrr` {#sec-purrr}

`purrr` provides a consistent, type-safe interface for functional programming:

```{r}
#| label: purrr-demo

# map: always returns a list
map(1:5, ~ .x^2)

# map_dbl: returns a double vector
map_dbl(mtcars, mean)

# map_df: returns a data frame
iris |>
  group_split(Species) |>
  map_dfr(~ tibble(
    species    = unique(.x$Species),
    mean_sepal = mean(.x$Sepal.Length),
    n          = nrow(.x)
  ))
```

```{r}
#| label: purrr-map2

# map2: iterate over two vectors simultaneously
x_vals <- 1:5
y_vals <- 6:10
map2_dbl(x_vals, y_vals, ~ .x + .y)

# pmap: iterate over any number of inputs
params <- list(
  mean = c(0, 5, 10),
  sd   = c(1, 2, 3),
  n    = c(100, 100, 100)
)
pmap(params, rnorm) |> map_dbl(mean)
```

```{r}
#| label: purrr-practical

# Practical: fit a model for each species, extract R-squared
iris |>
  group_split(Species) |>
  map_dfr(function(df) {
    m <- lm(Sepal.Length ~ Sepal.Width + Petal.Length, data = df)
    tibble(
      species   = unique(df$Species),
      r_squared = summary(m)$r.squared,
      n         = nrow(df)
    )
  })
```

## Vectorisation and Performance {#sec-performance}

```{r}
#| label: vectorisation

# Slow: loop over elements
sum_loop <- function(x) {
  total <- 0
  for (val in x) total <- total + val
  total
}

# Fast: vectorised
n <- 1e6
x <- rnorm(n)

bench::mark(
  loop   = sum_loop(x),
  base_r = sum(x),
  check  = FALSE
)
```

```{r}
#| label: memory-preallocation

# Anti-pattern: growing a vector in a loop
slow_grow <- function(n) {
  x <- c()
  for (i in 1:n) x <- c(x, i^2)
  x
}

# Better: pre-allocate
fast_prealloc <- function(n) {
  x <- numeric(n)
  for (i in 1:n) x[i] <- i^2
  x
}

# Best: fully vectorised
fastest <- function(n) (1:n)^2

bench::mark(
  slow   = slow_grow(1000),
  better = fast_prealloc(1000),
  best   = fastest(1000),
  check  = TRUE
)
```

## Exercises {#sec-ch15-exercises}

1. Write a function using a `for` loop that takes a data frame and returns the number of missing values in each column. Then write the same function using `sapply()`. Then using `purrr::map_int()`.

2. Use `purrr::map_dfr()` to fit a linear regression of `mpg ~ wt` separately for each number of cylinders in `mtcars`. Return a data frame with the intercept, slope, and R-squared for each group.

3. Benchmark three approaches to computing row means of a 1000×100 matrix: (a) a for loop, (b) `apply()`, (c) `rowMeans()`. Which is fastest?

4. Use `purrr::possibly()` to handle errors gracefully when applying a function that might fail on some inputs.

5. **Challenge:** Read 12 monthly data files (simulate them using `write_csv()`), process each one, and combine into a single data frame using `purrr::map_dfr()` and `here::here()`.
