# Best Practices and Reproducibility {#sec-best-practices}

```{r}
#| label: setup-ch18
#| include: false
library(tidyverse)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Organise a research project using a consistent directory structure
- Use `renv` to create a reproducible package environment
- Implement basic Git version control from within RStudio
- Write clean, readable R code following community conventions
- Apply the principles of a reproducible research workflow
:::

## Project-Oriented Workflows {#sec-projects}

The foundation of reproducible research is a **self-contained project**. Every file needed to reproduce an analysis should live in one directory, with no hard-coded paths to files elsewhere on your computer.

### Recommended Directory Structure {#sec-dir-structure}

```
my-research-project/
├── _quarto.yml          # (if a Quarto book/website)
├── my-project.Rproj     # RStudio project file
├── renv.lock            # Package snapshot (managed by renv)
├── README.md            # Project overview
│
├── data/
│   ├── raw/             # Original, immutable data — never modify
│   └── processed/       # Cleaned, derived data
│
├── R/                   # Reusable R functions
│   ├── utils.R
│   └── plotting.R
│
├── analysis/            # Analysis scripts or .qmd files
│   ├── 01-clean.qmd
│   ├── 02-eda.qmd
│   └── 03-modelling.qmd
│
├── output/
│   ├── figures/
│   ├── tables/
│   └── reports/
│
└── docs/                # Rendered output (for GitHub Pages)
```

::: {.callout-important}
## The Two Rules of Raw Data

1. **Never modify raw data.** Put it in `data/raw/` and treat it as read-only.
2. **Always document its provenance.** Add a `data/raw/README.md` recording where the data came from, when it was downloaded, and what it contains.
:::

### The `here` Package Revisited {#sec-here-revisited}

```{r}
#| label: here-revisited

library(here)

# Paths always relative to the .Rproj location
raw_data_path    <- here("data", "raw", "crop_yields_2023.csv")
clean_data_path  <- here("data", "processed", "crop_yields_clean.rds")
figure_path      <- here("output", "figures", "yield_trend.png")

# These paths work on any machine, any OS
cat(raw_data_path, "\n")
```

### Naming Conventions {#sec-naming}

Consistent file and variable naming pays dividends in readability:

```r
# Files: use lowercase with hyphens or underscores, be descriptive
# Good:
01-data-cleaning.qmd
02-eda-wheat-prices.qmd
fig-yield-by-state.png

# Bad:
analysis.R
Final_FINAL_v3.qmd
myfig.png

# Objects: use snake_case consistently
# Good:
wheat_yield_2023 <- read_csv(...)
mean_yield_state <- ...

# Bad:
WheatYield <- ...
meanyield.2023 <- ...
x <- ...
```

## Version Control with Git {#sec-git}

**Git** tracks every change to your files, allows you to revert to any previous state, and enables collaboration. **GitHub** hosts Git repositories online.

### Basic Git Workflow {#sec-git-workflow}

```bash
# Initialise a new repository
git init

# Check status
git status

# Stage files for commit
git add .                         # Stage all changed files
git add analysis/01-clean.qmd     # Stage one file

# Commit with a message
git commit -m "Add data cleaning script for wheat yields"

# View history
git log --oneline

# Push to GitHub
git remote add origin https://github.com/username/my-project.git
git push -u origin main
```

### Branching {#sec-branches}

```bash
# Create a new branch for an experiment
git checkout -b feature/add-poisson-model

# Make changes, commit...
git commit -m "Add Poisson regression for hospital admissions"

# Merge back to main
git checkout main
git merge feature/add-poisson-model
```

### Git in RStudio {#sec-git-rstudio}

RStudio provides a GUI for Git:

1. *Tools → Version Control → Project Setup* — enables Git for a project
2. The **Git pane** (Environment panel, Git tab) shows changed files
3. Click **Diff** to see exactly what changed
4. Check boxes to **Stage**, then click **Commit**
5. Use **Pull** before starting work; **Push** when done

::: {.callout-tip}
## What to Put in `.gitignore`

Some files should not be tracked. Create a `.gitignore` file:

```
# R artefacts
.Rhistory
.RData
.Rproj.user/

# Large data files (share via cloud or DOI instead)
data/raw/*.csv
data/raw/*.xlsx

# Rendered outputs (can be regenerated)
_book/
docs/
*.pdf

# renv package library (renv.lock is enough)
renv/library/
```
:::

## Package Management with `renv` {#sec-renv}

`renv` creates a **project-specific package library** and records exact package versions, so anyone (including future-you) can reproduce your exact environment.

```{r}
#| eval: false

library(renv)

# 1. Initialise renv in your project
renv::init()
# Creates: renv/ directory, renv.lock, .Rprofile

# 2. Install packages as usual
install.packages("gapminder")

# 3. Snapshot the current state
renv::snapshot()
# Updates renv.lock with all current package versions

# 4. On a new machine or after cloning from GitHub
renv::restore()
# Installs exactly the packages recorded in renv.lock
```

The `renv.lock` file looks like this:

```json
{
  "R": {
    "Version": "4.3.2",
    "Repositories": [{"Name": "CRAN", "URL": "https://cloud.r-project.org"}]
  },
  "Packages": {
    "ggplot2": {
      "Package": "ggplot2",
      "Version": "3.4.4",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "..."
    }
  }
}
```

::: {.callout-tip}
## Commit `renv.lock`, not `renv/library/`

Add `renv.lock` to Git to share the package snapshot. Add `renv/library/` to `.gitignore` — it is large and platform-specific.
:::

## Writing Clean R Code {#sec-clean-code}

### The tidyverse Style Guide {#sec-style}

```{r}
#| eval: false

# Spacing: spaces around operators and after commas
x <- 1 + 2          # Good
x<-1+2              # Bad

# Line length: max ~80 characters; break long pipes
result <- my_data |>
  filter(year == 2023) |>
  group_by(state) |>
  summarise(mean_yield = mean(yield, na.rm = TRUE))

# Not:
result <- my_data |> filter(year == 2023) |> group_by(state) |> summarise(mean_yield = mean(yield, na.rm = TRUE))

# Comments: explain WHY, not WHAT
# Compute yield index relative to national average
yield_index <- state_yield / national_avg_yield   # Good comment

x <- state_yield / national_avg_yield   # divide state by national avg   # Bad comment

# Function naming: use verbs
compute_yield_index()   # Good
yield_index_function()  # Bad
yi()                    # Bad
```

### Using `lintr` and `styler` {#sec-linting}

```{r}
#| eval: false

# lintr: check code style automatically
install.packages("lintr")
lintr::lint("analysis/01-clean.R")    # Check one file
lintr::lint_dir("R/")                 # Check all files in a directory

# styler: automatically reformat code to tidyverse style
install.packages("styler")
styler::style_file("analysis/01-clean.R")
styler::style_dir("analysis/")
```

## A Complete Reproducible Workflow {#sec-full-workflow}

Putting it all together, a fully reproducible analysis follows this sequence:

```{r}
#| eval: false
#| label: full-workflow

# 1. PROJECT SETUP (once)
# - Create RStudio Project
# - Initialise Git: git init
# - Initialise renv: renv::init()
# - Create directory structure

# 2. DATA ACQUISITION (data/raw/ — never modified)
# - Download/receive raw data
# - Document provenance in data/raw/README.md
# - Commit: git commit -m "Add raw data files"

# 3. DATA CLEANING (analysis/01-clean.qmd)
library(tidyverse)
library(here)
library(janitor)

raw <- read_csv(here("data", "raw", "crop_yields_raw.csv"),
                na = c("", "NA", "N/A", "--"))
clean <- raw |>
  clean_names() |>
  filter(!is.na(state)) |>
  mutate(year = as.integer(year))

saveRDS(clean, here("data", "processed", "crop_yields_clean.rds"))

# 4. ANALYSIS (analysis/02-eda.qmd, 03-model.qmd, ...)
# - Load from data/processed/
# - All outputs go to output/

# 5. SNAPSHOT AND COMMIT
# renv::snapshot()
# git add .
# git commit -m "Add EDA chapter with yield distributions"

# 6. PUBLISH
# quarto render
# quarto publish gh-pages
```

## The Turing Test for Reproducibility {#sec-turing-test}

A useful heuristic: **could a stranger reproduce your analysis from scratch?** More concretely:

1. Clone your GitHub repository onto a fresh computer
2. Run `renv::restore()` to install packages
3. Run `quarto render` or execute scripts in order
4. Do you get identical results?

If yes, your work is reproducible. If no, something is missing.

::: {.callout-important}
## The Three Enemies of Reproducibility

1. **Hard-coded paths**: `read_csv("C:/Users/pawan/Desktop/data.csv")` — use `here()` instead
2. **Random numbers without `set.seed()`**: results change every run
3. **Undocumented package versions**: `install.packages("ggplot2")` installs the *current* version, which changes over time — use `renv`
:::

## Exercises {#sec-ch18-exercises}

1. Create a new RStudio Project with the recommended directory structure. Initialise `renv` and install three packages. Take a snapshot. List what is recorded in `renv.lock`.

2. Initialise Git in your project. Make three commits — one for the project structure, one after adding a data file, and one after writing an analysis script. View the log with `git log --oneline`.

3. Apply `lintr::lint_dir()` to any existing R script you have written. Fix at least three style issues it identifies.

4. Write a `README.md` for a hypothetical project. It should include: project title, objective, data sources, how to reproduce the analysis, and required R version.

5. **Challenge:** Take any analysis you have done in a previous chapter and make it fully reproducible: clean directory structure, `renv.lock`, `.gitignore`, parameterised Quarto report, and published to GitHub. Share the URL.
