# Data Manipulation with dplyr and tidyr {#sec-dplyr}

```{r}
#| label: setup-ch4
#| include: false
library(tidyverse)
```

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

- Understand the philosophy of tidy data
- Use the six core `dplyr` verbs fluently
- Chain operations with the pipe operator (`|>`)
- Perform grouped summaries with `group_by()` and `summarise()`
- Reshape data between wide and long formats with `tidyr`
- Use `data.table` for high-performance operations on large datasets
:::

## The Tidyverse Philosophy: Tidy Data {#sec-tidy-data}

**Tidy data** is a consistent way of organising tabular data that makes analysis much easier. Hadley Wickham formalised the concept in his 2014 paper. The three rules of tidy data are:

1. Each **variable** forms a **column**
2. Each **observation** forms a **row**
3. Each **type of observational unit** forms a **table**

Most real-world data violates at least one of these rules. The tidyverse is designed to make it easy to tidy messy data.

```{r}
#| label: tidy-vs-messy

# Messy: crop yields spread across year columns
messy_yield <- tibble(
  state   = c("Punjab", "Haryana", "UP"),
  crop    = c("Wheat", "Wheat", "Rice"),
  `2021`  = c(48.2, 45.1, 30.5),
  `2022`  = c(49.5, 46.3, 31.2),
  `2023`  = c(50.1, 47.8, 32.1)
)
print(messy_yield)

# Tidy: each row is one observation (state × crop × year)
tidy_yield <- messy_yield |>
  pivot_longer(
    cols      = `2021`:`2023`,
    names_to  = "year",
    values_to = "yield_qtl_ha"
  ) |>
  mutate(year = as.integer(year))
print(tidy_yield)
```

## The Pipe Operator {#sec-pipe}

The **pipe operator** `|>` (base R, version 4.1+) passes the result on its left as the first argument to the function on its right. This enables readable, left-to-right chains of operations.

```{r}
#| label: pipe-demo

# Without pipe: nested, reads inside-out
result <- round(mean(sqrt(c(4, 9, 16, 25))), 2)

# With pipe: linear, reads left-to-right
result <- c(4, 9, 16, 25) |>
  sqrt() |>
  mean() |>
  round(2)

print(result)
```

::: {.callout-tip}
## `|>` vs `%>%`

The `|>` is the native R pipe introduced in R 4.1.0. The `%>%` is the magrittr pipe (from the `magrittr` package, loaded with `dplyr`). They are nearly identical for most uses. This book uses `|>` throughout to avoid loading an extra dependency.

The main difference: `|>` always passes to the *first* argument; `%>%` allows using `.` as a placeholder for other positions.
:::

## Core dplyr Verbs {#sec-dplyr-verbs}

`dplyr` provides **six core verbs** that cover the vast majority of data manipulation needs.

### `filter()` — Subset Rows {#sec-filter}

```{r}
#| label: filter-demo

data(gapminder, package = "gapminder")

# Single condition
india <- gapminder |>
  filter(country == "India")

# Multiple conditions (AND)
south_asia_2007 <- gapminder |>
  filter(continent == "Asia", year == 2007, lifeExp > 70)

# OR condition
india_china <- gapminder |>
  filter(country %in% c("India", "China"))

# Negation
not_africa <- gapminder |>
  filter(continent != "Africa")

# Filter with string matching
# filter(str_detect(country, "^I"))   # Countries starting with I

nrow(india)
nrow(south_asia_2007)
```

### `select()` — Subset Columns {#sec-select}

```{r}
#| label: select-demo

# Select by name
gapminder |>
  select(country, year, lifeExp) |>
  head(4)

# Select a range
gapminder |>
  select(country:lifeExp) |>
  head(4)

# Drop columns with -
gapminder |>
  select(-continent) |>
  head(4)

# Helper functions
gapminder |>
  select(starts_with("c")) |>
  head(4)

gapminder |>
  select(ends_with("p")) |>
  head(4)

gapminder |>
  select(where(is.numeric)) |>
  head(4)
```

### `mutate()` — Create or Modify Columns {#sec-mutate}

```{r}
#| label: mutate-demo

gapminder_ext <- gapminder |>
  mutate(
    gdp_total      = gdpPercap * pop,               # New column
    log_gdp        = log(gdpPercap),                # Log transformation
    lifeExp_decade = lifeExp %/% 10 * 10,           # Decade bins
    high_income    = gdpPercap > median(gdpPercap), # Logical column
    pop_millions   = pop / 1e6                      # Rescale
  )

gapminder_ext |>
  select(country, year, gdpPercap, gdp_total, high_income) |>
  head(5)
```

### `arrange()` — Sort Rows {#sec-arrange}

```{r}
#| label: arrange-demo

# Ascending (default)
gapminder |>
  filter(year == 2007) |>
  arrange(lifeExp) |>
  head(5)

# Descending
gapminder |>
  filter(year == 2007) |>
  arrange(desc(gdpPercap)) |>
  head(5)

# Multiple sort keys
gapminder |>
  arrange(continent, desc(year), lifeExp) |>
  head(5)
```

### `summarise()` — Compute Summary Statistics {#sec-summarise}

```{r}
#| label: summarise-demo

gapminder |>
  filter(year == 2007) |>
  summarise(
    n_countries    = n(),
    mean_lifeExp   = mean(lifeExp),
    median_lifeExp = median(lifeExp),
    sd_lifeExp     = sd(lifeExp),
    min_lifeExp    = min(lifeExp),
    max_lifeExp    = max(lifeExp)
  )
```

### `group_by()` + `summarise()` — Grouped Summaries {#sec-group-by}

This combination is the workhorse of data analysis — compute statistics *within* groups:

```{r}
#| label: group-by-demo

gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    n_countries  = n(),
    mean_gdp     = mean(gdpPercap),
    mean_lifeExp = mean(lifeExp),
    total_pop    = sum(pop)
  ) |>
  arrange(desc(mean_lifeExp))
```

```{r}
#| label: group-mutate-demo

# group_by + mutate: compute relative to group (not summarise)
gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  mutate(
    gdp_rel_to_continent = gdpPercap / mean(gdpPercap),
    gdp_rank_in_continent = rank(desc(gdpPercap))
  ) |>
  filter(country %in% c("India", "China", "Japan")) |>
  select(country, continent, gdpPercap, gdp_rel_to_continent, gdp_rank_in_continent)
```

### `rename()` and `relocate()` {#sec-rename-relocate}

```{r}
#| label: rename-demo

gapminder |>
  rename(
    gdp_per_capita = gdpPercap,
    life_expectancy = lifeExp,
    population = pop
  ) |>
  relocate(population, .before = life_expectancy) |>
  head(3)
```

## Joining Data Frames {#sec-joins}

Real analysis almost always involves combining multiple tables. `dplyr` provides a family of **join** functions:

```{r}
#| label: joins-demo

# Example: join crop yields with state population data
crop_yield <- tibble(
  state  = c("Punjab", "Haryana", "UP", "MP"),
  crop   = c("Wheat", "Wheat", "Rice", "Soybean"),
  yield  = c(50.2, 47.8, 32.1, 12.5)
)

state_pop <- tibble(
  state      = c("Punjab", "Haryana", "UP", "Rajasthan"),
  population = c(27.7, 25.4, 199.8, 68.5)  # millions
)

# Inner join: only rows with matches in BOTH tables
inner_join(crop_yield, state_pop, by = "state")

# Left join: all rows from crop_yield, NA for non-matches
left_join(crop_yield, state_pop, by = "state")

# Full join: all rows from both tables
full_join(crop_yield, state_pop, by = "state")
```

## Reshaping Data with `tidyr` {#sec-tidyr}

### `pivot_longer()` — Wide to Long {#sec-pivot-longer}

```{r}
#| label: pivot-longer-demo

# Monthly rainfall data (wide format — common in reports)
rainfall_wide <- tibble(
  station = c("Ludhiana", "Chandigarh", "Amritsar"),
  Jan = c(28, 32, 25),
  Feb = c(35, 38, 30),
  Mar = c(22, 19, 18),
  Apr = c(8, 7, 6)
)

print(rainfall_wide)

# Convert to long (tidy) format
rainfall_long <- rainfall_wide |>
  pivot_longer(
    cols      = Jan:Apr,
    names_to  = "month",
    values_to = "rainfall_mm"
  )

print(rainfall_long)
```

### `pivot_wider()` — Long to Wide {#sec-pivot-wider}

```{r}
#| label: pivot-wider-demo

# Convert back to wide (useful for creating summary tables)
rainfall_long |>
  pivot_wider(
    names_from  = month,
    values_from = rainfall_mm
  )

# A more complex example: multiple variables
panel_long <- tibble(
  state = rep(c("Punjab", "Haryana"), each = 3),
  year  = rep(2021:2023, 2),
  gdp   = c(6.5, 6.8, 7.1, 5.2, 5.4, 5.7),
  pop   = c(27.5, 27.7, 27.9, 25.2, 25.4, 25.6)
)

panel_wide <- panel_long |>
  pivot_wider(
    names_from  = year,
    values_from = c(gdp, pop),
    names_sep   = "_"
  )

print(panel_wide)
```

### Handling Missing Values {#sec-missing}

```{r}
#| label: missing-demo

df_with_na <- tibble(
  x = c(1, NA, 3, NA, 5),
  y = c("a", "b", NA, "d", "e")
)

# Detect missing
is.na(df_with_na$x)
sum(is.na(df_with_na$x))   # Count NAs

# Drop rows with any NA
df_with_na |> drop_na()

# Drop rows with NA in specific columns
df_with_na |> drop_na(x)

# Fill NAs
df_with_na |>
  mutate(x_filled = replace_na(x, 0))

# Fill forward/backward
df_with_na |>
  mutate(x_forward = tidyr::fill(pick(x), x)$x)
```

## `data.table` for Large Datasets {#sec-datatable}

For datasets with millions of rows, `data.table` offers dramatically faster performance than `dplyr`.

```{r}
#| label: datatable-demo
library(data.table)

# Convert to data.table
mtcars_dt <- as.data.table(mtcars)

# Syntax: DT[i, j, by]
# i = row filter, j = column operations, by = grouping

# Filter rows
mtcars_dt[cyl == 6]

# Select columns
mtcars_dt[, .(mpg, cyl, hp)]

# Add/modify column
mtcars_dt[, mpg_per_hp := mpg / hp]

# Group by and summarise
mtcars_dt[, .(mean_mpg = mean(mpg), count = .N), by = cyl]
```

::: {.callout-tip}
## When to Use `data.table`

- Datasets with **>1 million rows**: `data.table` is 5–50× faster than `dplyr`
- Memory-constrained environments: `data.table` modifies objects in-place
- Production pipelines requiring maximum speed

For typical research datasets (<100,000 rows), `dplyr`'s more readable syntax is preferred.
:::

## A Complete Data Wrangling Pipeline {#sec-pipeline}

```{r}
#| label: full-pipeline

# Simulate a realistic messy dataset
set.seed(42)
raw_data <- tibble(
  district   = rep(c("Ludhiana", "Amritsar", "Jalandhar", "Patiala"), each = 5),
  year       = rep(2019:2023, 4),
  wheat_prod = round(runif(20, 800, 1500)),
  rice_prod  = round(runif(20, 200, 600)),
  area_ha    = round(runif(20, 100, 300)),
  notes      = sample(c("verified", "estimated", NA), 20, replace = TRUE)
)

# Full wrangling pipeline
clean_data <- raw_data |>
  # 1. Filter recent years only
  filter(year >= 2021) |>
  # 2. Handle missing notes
  mutate(notes = replace_na(notes, "unknown")) |>
  # 3. Create derived variables
  mutate(
    total_prod      = wheat_prod + rice_prod,
    productivity    = total_prod / area_ha,
    high_yield      = productivity > median(productivity)
  ) |>
  # 4. Reshape to long format
  pivot_longer(
    cols      = ends_with("_prod"),
    names_to  = "crop",
    values_to = "production"
  ) |>
  mutate(crop = str_remove(crop, "_prod")) |>
  # 5. Summarise by district and crop
  group_by(district, crop) |>
  summarise(
    mean_production = round(mean(production), 1),
    total_years     = n(),
    .groups         = "drop"
  ) |>
  # 6. Sort
  arrange(crop, desc(mean_production))

print(clean_data)
```

## Exercises {#sec-ch4-exercises}

1. Using the `gapminder` dataset, find the five countries with the highest GDP per capita in 2007. Show only the country name, continent, and GDP per capita.

2. Compute the average life expectancy per continent for each decade (1950s, 1960s, ..., 2000s). *Hint:* Create a decade variable using integer division.

3. The following dataset is in wide format. Convert it to tidy (long) format, then compute the mean score per student:

```r
scores <- tibble(
  student = c("Aarav", "Bhavna", "Chirag"),
  maths   = c(85, 92, 78),
  english = c(72, 88, 95),
  science = c(90, 76, 83)
)
```

4. Join the `crop_yield` and `state_pop` tibbles from @sec-joins using a `right_join()`. What is the result? How does it differ from `left_join()`?

5. **Challenge:** Download state-wise crop production data from data.gov.in. Import, clean (handle NAs, standardise column names), reshape to tidy format, and compute the top 5 producing states for each crop in the most recent year available.
