# Appendix C: Common R Packages for Data Science {#sec-appendix-c .unnumbered}

This appendix catalogs the R packages used throughout this book, organised by task, with brief descriptions and installation commands.

## Installing All Book Packages {#sec-install-all .unnumbered}

```r
# Run this once to install every package used in this book
pkgs <- c(
  # Core tidyverse
  "tidyverse", "tibble", "dplyr", "tidyr", "readr",
  "ggplot2", "purrr", "stringr", "forcats", "lubridate",

  # Data import
  "readxl", "haven", "jsonlite", "googlesheets4",
  "writexl", "DBI", "RSQLite", "RPostgres",

  # Data cleaning
  "janitor", "skimr", "naniar",

  # Visualisation
  "plotly", "corrplot", "GGally", "ggfortify",
  "patchwork", "scales", "RColorBrewer", "viridis",
  "ggthemes", "ggrepel",

  # Statistics
  "moments", "effectsize", "car", "multcomp", "emmeans",

  # Regression and modelling
  "broom", "modelsummary", "gtsummary", "gt",
  "lme4", "broom.mixed", "MASS", "performance",

  # Multivariate
  "factoextra", "cluster", "FactoMineR",

  # Time series
  "forecast", "tseries", "tsibble", "feasts", "fable",

  # Machine learning
  "caret", "tidymodels",

  # Functional programming / utilities
  "purrr", "furrr", "bench", "microbenchmark",
  "here", "renv", "fs", "glue",

  # Quarto / reporting
  "knitr", "rmarkdown", "quarto",

  # Debugging / style
  "lintr", "styler",

  # Datasets
  "gapminder", "palmerpenguins", "nycflights13"
)

# Install packages not already installed
new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) install.packages(new_pkgs)
```

---

## Package Catalogue {#sec-pkg-catalogue .unnumbered}

### Data Import and Export {#sec-pkg-import .unnumbered}

```{r}
#| label: tbl-import-pkgs
#| echo: false
#| tbl-cap: "Packages for data import and export."

import_pkgs <- data.frame(
  Package = c("readr", "readxl", "haven", "writexl", "jsonlite",
              "DBI", "RSQLite", "RPostgres", "googlesheets4", "httr2"),
  Function = c(
    "Read CSV, TSV, and delimited text files",
    "Read Excel files (.xls, .xlsx) without Java",
    "Read SPSS (.sav), Stata (.dta), SAS (.sas7bdat) files",
    "Write data frames to Excel without Java",
    "Parse and generate JSON",
    "Unified database interface",
    "Interface to SQLite databases",
    "Interface to PostgreSQL databases",
    "Read and write Google Sheets",
    "HTTP requests for web APIs"
  ),
  Install = c(
    "tidyverse", "tidyverse", "install.packages('haven')",
    "install.packages('writexl')", "install.packages('jsonlite')",
    "install.packages('DBI')", "install.packages('RSQLite')",
    "install.packages('RPostgres')", "install.packages('googlesheets4')",
    "install.packages('httr2')"
  )
)
knitr::kable(import_pkgs, col.names = c("Package", "Purpose", "Install"))
```

### Data Wrangling {#sec-pkg-wrangle .unnumbered}

```{r}
#| label: tbl-wrangle-pkgs
#| echo: false
#| tbl-cap: "Packages for data manipulation and cleaning."

wrangle_pkgs <- data.frame(
  Package = c("dplyr", "tidyr", "data.table", "janitor", "skimr",
              "lubridate", "stringr", "forcats", "naniar", "validate"),
  Purpose = c(
    "Core data manipulation: filter, select, mutate, summarise",
    "Reshape data: pivot_longer, pivot_wider, separate, unite",
    "High-performance data manipulation for large datasets",
    "Clean data frame names and values (clean_names)",
    "Rich summary statistics with skim()",
    "Parse, manipulate, and arithmetic on dates and times",
    "Consistent string manipulation functions",
    "Tools for working with categorical (factor) variables",
    "Visualise and analyse missing data patterns",
    "Validate data against rules"
  )
)
knitr::kable(wrangle_pkgs, col.names = c("Package", "Purpose"))
```

### Visualisation {#sec-pkg-viz .unnumbered}

```{r}
#| label: tbl-viz-pkgs
#| echo: false
#| tbl-cap: "Packages for data visualisation."

viz_pkgs <- data.frame(
  Package = c("ggplot2", "plotly", "patchwork", "corrplot",
              "GGally", "ggrepel", "ggthemes", "viridis",
              "scales", "ggfortify", "leaflet"),
  Purpose = c(
    "Grammar of Graphics — the core visualisation package",
    "Interactive charts; converts ggplot2 with ggplotly()",
    "Combine multiple ggplot2 plots with / and |",
    "Visualise correlation matrices",
    "Pairs plots, parallel coordinates, and ggplot2 extensions",
    "Non-overlapping text labels for scatter plots",
    "Additional themes (Economist, Tufte, FiveThirtyEight, ...)",
    "Perceptually uniform colour palettes (colourblind-safe)",
    "Format axis labels (dollar, percent, comma, ...)",
    "autoplot() for time series, PCA, survival objects",
    "Interactive choropleth and point maps"
  )
)
knitr::kable(viz_pkgs, col.names = c("Package", "Purpose"))
```

### Statistical Modelling {#sec-pkg-models .unnumbered}

```{r}
#| label: tbl-model-pkgs
#| echo: false
#| tbl-cap: "Packages for statistical modelling and inference."

model_pkgs <- data.frame(
  Package = c("stats", "car", "multcomp", "emmeans", "lme4",
              "MASS", "survival", "nlme", "mgcv", "glmnet",
              "broom", "modelsummary", "gtsummary", "effectsize",
              "performance", "see"),
  Purpose = c(
    "Base R statistics: lm, glm, t.test, aov, etc.",
    "Companion to Applied Regression: ANOVA, VIF, leveneTest",
    "Multiple comparisons and simultaneous inference",
    "Estimated marginal means for factorial designs",
    "Linear and generalised linear mixed models",
    "Negative binomial regression, stepwise selection, LDA",
    "Survival analysis: Kaplan-Meier, Cox regression",
    "Mixed effects models with correlation structures",
    "Generalised additive models (GAMs)",
    "LASSO, Ridge, and Elastic Net regression",
    "Tidy model outputs: tidy, glance, augment",
    "Publication-quality regression tables",
    "Clinical summary tables with gtsummary",
    "Standardised effect sizes (Cohen's d, eta-squared, ...)",
    "Model quality indices: R², AIC, ICC, RMSE",
    "Visualisation for easystats packages"
  )
)
knitr::kable(model_pkgs, col.names = c("Package", "Purpose"))
```

### Time Series {#sec-pkg-ts .unnumbered}

```{r}
#| label: tbl-ts-pkgs
#| echo: false
#| tbl-cap: "Packages for time series analysis and forecasting."

ts_pkgs <- data.frame(
  Package = c("forecast", "tseries", "tsibble", "feasts",
              "fable", "prophet", "xts", "zoo"),
  Purpose = c(
    "ARIMA, ETS, Holt-Winters forecasting; auto.arima()",
    "Unit root tests (adf.test), ARCH effects",
    "Tidy time series data structure",
    "Feature extraction and visualisation for tsibble",
    "Tidy forecasting framework (ARIMA, ETS, ...)",
    "Facebook's additive forecasting model (trends + seasonality)",
    "Extensible time series for irregular data",
    "Ordered observations for irregular time series"
  )
)
knitr::kable(ts_pkgs, col.names = c("Package", "Purpose"))
```

### Reproducibility and Workflow {#sec-pkg-repro .unnumbered}

```{r}
#| label: tbl-repro-pkgs
#| echo: false
#| tbl-cap: "Packages for reproducible research workflows."

repro_pkgs <- data.frame(
  Package = c("renv", "here", "targets", "quarto", "knitr",
              "rmarkdown", "lintr", "styler", "usethis", "devtools"),
  Purpose = c(
    "Reproducible package environments with lockfile",
    "File paths relative to project root",
    "Make-style pipeline for complex, cached workflows",
    "Render Quarto documents from R",
    "R code chunks in documents; kable() for tables",
    "Dynamic documents with R Markdown",
    "Static code analysis and style checking",
    "Automatic code reformatting to tidyverse style",
    "Automate project and package setup tasks",
    "Package development tools"
  )
)
knitr::kable(repro_pkgs, col.names = c("Package", "Purpose"))
```

### Datasets {#sec-pkg-data .unnumbered}

```{r}
#| label: tbl-data-pkgs
#| echo: false
#| tbl-cap: "Packages that provide datasets for learning and examples."

data_pkgs <- data.frame(
  Package = c("gapminder", "palmerpenguins", "nycflights13",
              "ISLR2", "wooldridge", "AER", "datasets"),
  Key_Datasets = c(
    "gapminder: GDP, life expectancy, population for 142 countries, 1952–2007",
    "penguins: measurements for 344 penguins of 3 species",
    "flights, airlines, airports, planes, weather (2013)",
    "Auto, Boston, Caravan, Wage, and many others (ISLR textbook)",
    "100+ datasets for Wooldridge's Econometrics textbook",
    "CPS1985, Fatalities, PSID, and many applied econometrics datasets",
    "mtcars, iris, airquality, USArrests, etc. (built into R)"
  )
)
knitr::kable(data_pkgs, col.names = c("Package", "Key Datasets"))
```

---

## Package Discovery {#sec-pkg-discovery .unnumbered}

Finding the right package for a new task:

- **CRAN Task Views**: [cran.r-project.org/web/views/](https://cran.r-project.org/web/views/) — curated lists by topic (Econometrics, TimeSeries, Spatial, ...)
- **rOpenSci**: [ropensci.org](https://ropensci.org) — peer-reviewed packages for scientific data
- **Posit Community**: [community.rstudio.com](https://community.rstudio.com)
- **R-bloggers**: [r-bloggers.com](https://www.r-bloggers.com)
- **Stack Overflow**: tag `[r]` for questions

---

## Getting Package Help {#sec-pkg-help .unnumbered}

```r
# Built-in help
?dplyr::filter
help(package = "ggplot2")
vignette("dplyr")                  # Package tutorial
vignette(package = "tidyr")        # List all vignettes

# Check package version
packageVersion("ggplot2")

# List all functions in a package
ls("package:dplyr")

# Package news (changelog)
news(package = "ggplot2")
```
